---
title: "ex7_heeju"
author: "Heeju Noh"
date: "November 17, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### **Exercise 7**


```{r eval=TRUE, warning=FALSE, message=FALSE}
library("limma")
library("affy")
library("preprocessCore")
unzip("affy_estrogen.zip")
ddir <- "C:/Users/Heeju/Desktop/STAT426/affy_estrogen"
dir(ddir)
```


##### **Preprocess affymetrix data**
```{r eval=TRUE, warning=FALSE, message=FALSE}
targets <- readTargets("targets.txt", path=ddir)
targets

abatch <- ReadAffy(filenames=targets$filename,
          celfile.path=ddir)
eset <- rma(abatch)  # bg correct, normalize, summarize
```



##### **Multidimensional scailing (MDS) plot**
```{r eval=TRUE, warning=FALSE, message=FALSE}
plotMDS(exprs(eset) )  # MDS plot
```


##### **Design matrix**
```{r}
# do the limma modeling
f <- paste(targets$estrogen,targets$time.h,sep="")
f <- factor(f)

# create design matrix
design <- model.matrix(~0+f)
colnames(design) <- levels(f)
design
fit <- lmFit(eset, design)
```


##### **Contrast matrix**
```{r}
cont.matrix <- makeContrasts(E10="present10-absent10",
                             E48="present48-absent48",
                             Time="absent48-absent10",levels=design)
cont.matrix
fit2  <- contrasts.fit(fit, cont.matrix)
fit2  <- eBayes(fit2)

class(fit2)
names(fit2)
```


##### **Differential expression statistics**
```{r}
topTable(fit2,coef=1)
topTable(fit2,coef=2)
barplot( exprs(eset)["39642_at",], las=2, cex.names=.7 )  # top gene
```


##### **Question 1.  From the matrix of summarized Affymetrix data that went into the limma pipeline in the first place -- exprs(eset) -- manually calculate the logFC and AveExpr for one of the top differentially expressed genes.**

From the topTables, I chose "910_at" as one of the top differentially expressed genes, and calculated logFC and average expressions over samples manually. The results of AveEXP and logFC were consistent with the values in topTables.

```{r}
probeID <- which(rownames(exprs(eset))=="910_at")
exp_condi <- (t(design) %*% exprs(eset)[probeID,])/2
logFC <- t(cont.matrix) %*% exp_condi
AveEXP <- mean(exp_condi)

AveEXP
logFC
```



##### **Question 2. Venn diagram of differentially expressed genes among E10, E48 and Time.**
```{r}
a <- decideTests(fit2) # default setting: p-value=0.05, lfc=0, adjust.method="BH"
vennDiagram(a)
```


##### **Question 3. Try to reproduce the above limma modeling using an alternative design matrix**

I set a new design matrix as depicted below and calculated the coefficients of the linear model:

```{r}
# do the limma modeling
f1 <- targets$estrogen
f1 <- factor(f1)
f2 <- targets$time.h
f2 <- factor(f2)

# create design matrix
design2 <- model.matrix(~f1+f2+f1:f2)
colnames(design2) <- c("a1","a2","a3","a4")
design2[7:8,2]<-0
design2
fit <- lmFit(eset, design2)
fit  <- eBayes(fit)
```

Now in this model, the expressions can be calculated as described as below:

**absent10** = a1,
**present10** = a1+a2,
**absent48** = a1+a3,
**present48** = a1+a3+a4,

**E10 = "present10-absent10"** = a2,
**E48 = "present48-absent48"** = a4,
**Time = "absent48-absent10"** = a3,

where a1, a2, a3, and a4 are the coefficients of the linear model.
The results from new design matrix were the same as those from the first way, and the same statistical results could be reproduced. 

```{r}
topTable(fit,coef=2) # E10
topTable(fit,coef=4) # E48
b <- decideTests(fit[,2:4]) # default setting: p-value=0.05, lfc=0, adjust.method="BH"
colnames(b) <- c("E10","Time","E48")
vennDiagram(b)
```